Modal Abstractions for Virtualizing Memory Addresses

### 2023-03-05 programlogic.tex
+

\subsection{Address-Space Abstraction}
\label{sec:aspacemodal}
Abstractions we introduce for virtual address spaces are based on the following aspects:
\begin{enumerate}
\item a global address space invariant asserting the authorative ownership of the ghost map, $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$ 
\item for each ghost page-table-walk mapping,  $\sumwalkabs\vaddr\qfrac\paddr$, there exists concrete physical page-table accesses path
  \[\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\entryf,\entrytr,\entrytw,\entryo)\]
  \item all virtual pointsto assertions happen to be true in a certain address space, i.e. the truth on address virtualization is indexed by CPUs page table register ($\crthree$ for x86-64), and we present modal reasoning principles for updating $\crthree$ (switching from the current address space to another)  
\end{enumerate}
    \subsubsection{Global Per-Address-Space Invariant}
    \label{sec:peraspaceinvariant}
    Virtual points-to assertions would need to be defined in terms of physical points-to assertions and hardware-specific address translation assertions for mapped virtual addresses ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\entryf,\entrytr,\entrytw,\entryo)$ in Figure \ref{fig:peraspaceinvariant}). For each virtual address \textit{mapped} with a physical page table entry, abstracted with ghost mappings ($\sumwalkabs\vaddr\qfrac\paddr$) in virtul-pointsto relations as shown in Figure \ref{fig:virtualpointstosharing}), there exsits a physical page table walk.  
    \[ \bigast{(\vaddr, \entryo)\in \ptablestore}{\exists\;(\entryf\;,\entrytr\;,\entrytw)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\vaddr,\entryf,\entrytr,\entrytw,\entryo)} \]
    By separating the virtual-points Overlapping address spaces — both shared data, and shared page tables
which may see updates — can be coordinated with protocols to ensure those updates preserve
address space invariants.
\begin{figure*}
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace} \stackrel{\triangle}{=} \lambda \; \crthree \ldotp
  \exists\;\gammaPred \;\ldotp\; \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \\
  \bigast{(\vaddr, \entryo)\in \ptablestore}{\exists\;(\entryf\;,\entrytr\;,\entrytw)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\vaddr,\entryf,\entrytr,\entrytw,\entryo)}
\end{array}
\]
\caption{Global Address-Space Invariant}
  \label{fig:peraspaceinvariant}
\end{figure*}
\subsubsection{Address-Space Modality}
\label{sec:aspacemodalist}
The truth on an address space, exhibits itself as a contingent truth: a location virtualization assertion (a virtual points-to in \ref{fig:virtualpointstosharing}),  happens to valid in the \textit{current world} (in the current address space), and switching address spaces pulls information out of one \textit{world} into the “current view” of memory, and leaves other assertions true relative to the previous address space.

Therefore, an address-space, as an abstraction, can be treated as naming the memory state as a modal frame and the choice of page table root as a world in Kripke-style semantics. Not suprisingly, transition between two address-spaces, then, can be just an entailment relation \textit{alternating} the \textit{named-state} against the address-space invariant ($\mathcal{I}\textsf{ASpace}$ in Figure \ref{fig:peraspaceinvariant}). 
\[
  \begin{array}{l}
    \hbox{(\TirNameStyle{ModalAddressSpace})} \\ \qquad
         [r](P)  \;\stackrel{\triangle}{=}  (\mathcal{W}_{64} -\textsf{b}> \textsf{uPredI} (\textsf{iResUR} \;\Sigma)) -\textsf{b}> \textsf{uPredI} \_ := \lambda \textsf{P},\; \textsf{P r}. 
  \end{array}
\]
Being inspired by what hybrid logic \ref{} calls a satisfaction operator, which evaluates the truth of a predicate in a named alternative state (here, address space), we give a modal definition describing the truth of assertions for the resources (i.e. virtual pointsto relations) inside the address space. Ignoring the predicate types for a while, $[r]P$ in Definition $\hbox{(\TirNameStyle{ModalAddressSpace})}$, indicates that $P$ holds in the virtual address space rooted at $r$, the truth ($P$) on an address-space is indexed by the root page-table address $r$ of the address-space. In the rest of this section, we explain the structural aspects -- the modal resource context of address space modality -- , and how we lift the interaction of address-space modality with the ambient logic, i.e. separation logic as an entailment for specifying \textit{address-space-switch}.  
\paragraph{Modal Resource Context}
\label{sec:resourcecontext}
When we pick up a certain contingency, which is \textit{"a certain virtual-pointsto facts whose validity is indexed by the root address of an address space"}for address space modality, we, in fact, make our choice around some logical resources for which we care about reasoning, which is virtual pointsto relation shown in Figure \ref{fig:virtualpointstosharing}. In other words, our modality definition is the logical virtual pointsto resources: introduction/elimination of logical virtual pointsto resources are the ones essential to the reasoning capabilities the address space modality provides, i.e. validity of virtual addresses under/out of modality -- resources under/out of \textit{modal resource context}.

>>>>

\subsection{Address-Space Abstraction}
\label{sec:aspacemodal}
Abstractions we introduce for virtual address spaces are based on the following aspects:
\item a global address space invariant asserting the authorative ownership of the ghost map, $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$ 
\item for each ghost page-table-walk mapping,  $\sumwalkabs\vaddr\qfrac\paddr$, there exists concrete physical page-table accesses path
  \[\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\entryf,\entrytr,\entrytw,\entryo)\]
  \item all virtual pointsto assertions happen to be true in a certain address space, i.e. the truth on address virtualization is indexed by CPUs page table register ($\crthree$ for x86-64), and we present modal reasoning principles for updating $\crthree$ (switching from the current address space to another)  
    \subsubsection{Global Per-Address-Space Invariant}
    \label{sec:peraspaceinvariant}
    Virtual points-to assertions would need to be defined in terms of physical points-to assertions and hardware-specific address translation assertions for mapped virtual addresses ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\entryf,\entrytr,\entrytw,\entryo)$ in Figure \ref{fig:peraspaceinvariant}). For each virtual address \textit{mapped} with a physical page table entry, abstracted with ghost mappings ($\sumwalkabs\vaddr\qfrac\paddr$) in virtul-pointsto relations as shown in Figure \ref{fig:virtualpointstosharing}), there exsits a physical page table walk.  
    \[ \bigast{(\vaddr, \entryo)\in \ptablestore}{\exists\;(\entryf\;,\entrytr\;,\entrytw)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\vaddr,\entryf,\entrytr,\entrytw,\entryo)} \]
    By separating the virtual-points Overlapping address spaces — both shared data, and shared page tables
which may see updates — can be coordinated with protocols to ensure those updates preserve
address space invariants.
\begin{figure*}
\begin{array}{l}
  \mathcal{I}\textsf{ASpace} \stackrel{\triangle}{=} \lambda \; \crthree \ldotp
  \exists\;\gammaPred \;\ldotp\; \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \\
  \bigast{(\vaddr, \entryo)\in \ptablestore}{\exists\;(\entryf\;,\entrytr\;,\entrytw)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\vaddr,\entryf,\entrytr,\entrytw,\entryo)}
\caption{Global Address-Space Invariant}
  \label{fig:peraspaceinvariant}
\end{figure*}
\subsubsection{Address-Space Modality}
\label{sec:aspacemodalist}
The truth on an address space, exhibits itself as a contingent truth: a location virtualization assertion (a virtual points-to in \ref{fig:virtualpointstosharing}),  happens to valid in the \textit{current world} (in the current address space), and switching address spaces pulls information out of one \textit{world} into the “current view” of memory, and leaves other assertions true relative to the previous address space.
\begin{figure}
   \includegraphics[width=\columnwidth]{addr_space_switch.png}
  \caption{Loading an Address-Space into the Current Memory View}
  \label{fig:addrswitch}
Therefore, an address-space, as an abstraction, can be treated as naming the memory state as a modal frame and the choice of page table root as a world in Kripke-style semantics. Not suprisingly, transition between two address-spaces, then, can be just an entailment relation \textit{alternating} the \textit{named-state} against the address-space invariant ($\mathcal{I}\textsf{ASpace}$ in Figure \ref{fig:peraspaceinvariant}). 
  \begin{array}{l}
    \hbox{(\TirNameStyle{ModalAddressSpace})} \\ \qquad
         [r](P)  \;\stackrel{\triangle}{=}  (\mathcal{W}_{64} -\textsf{b}> \textsf{uPredI} (\textsf{iResUR} \;\Sigma)) -\textsf{b}> \textsf{uPredI} \_ := \lambda \textsf{P},\; \textsf{P r}. 
Being inspired by what hybrid logic \ref{} calls a satisfaction operator, which evaluates the truth of a predicate in a named alternative state (here, address space), we give a modal definition describing the truth of assertions for the resources (i.e. virtual pointsto relations) inside the address space. Ignoring the predicate types for a while, $[r]P$ in Definition $\hbox{(\TirNameStyle{ModalAddressSpace})}$, indicates that $P$ holds in the virtual address space rooted at $r$, the truth ($P$) on an address-space is indexed by the root page-table address $r$ of the address-space. In the rest of this section, we explain the structural aspects -- the modal resource context of address space modality -- , and how we lift the interaction of address-space modality with the ambient logic, i.e. separation logic as an entailment for specifying \textit{address-space-switch}.  
\paragraph{Modal Resource Context}
\label{sec:resourcecontext}
When we pick up a certain contingency, which is \textit{"a certain virtual-pointsto facts whose validity is indexed by the root address of an address space"}for address space modality, we, in fact, make our choice around some logical resources for which we care about reasoning, which is virtual pointsto relation shown in Figure \ref{fig:virtualpointstosharing}. In other words, our modality definition is the logical virtual pointsto resources: introduction/elimination of logical virtual pointsto resources are the ones essential to the reasoning capabilities the address space modality provides, i.e. validity of virtual addresses under/out of modality -- resources under/out of \textit{modal resource context}.
